---
title: "w11_async"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(car)
library(wooldridge)
```

The residual-fitted plot shows a slight increase in variance in the residuals, but the Q-Q plot confirms normality of the response with a few exceptions.  This is confirmed by the residuals-leverage plot which shows a few high leverage points.

```{r}
# summary(crime2)

crimes.82 = crime2[crime2$year==82,]
ols.fit1 = lm(crmrte ~ unem, data = crimes.82)
summary(ols.fit1)
plot(ols.fit1)
```

The residual-fitted plot shows that the residuals are heteroskedastic, and the Q-Q plot shows that the response is not a normal random variable.  The residuals-leverage plot which shows a few residuals with high leverage.

The negative coefficient for unemployment does not make sense as we would expect unemployment and crime rates to be positively correlated.

```{r}
# summary(crime2)

crimes.87 = crime2[crime2$year==87,]
ols.fit2 = lm(crmrte ~ unem, data = crimes.87)
summary(ols.fit2)
plot(ols.fit2)
```